SDGs

### 2023-06-14 inequality-SDGs.Rmd
national level analysis + wii new dataset

countries_list <- countries |> filter(!is.na(currency_unit)) |> pull(country_code)

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Year", y = "Indicator name")
    theme_light(base_size = 6)

@@

@@

@@

```{r}
b <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
     scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG14: Life under water", x = "Year", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6)
```

@@

## Very long names:
sdg|> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    select(indicator, indicator_name) |> 
    unique()

@@

@@

```{r}
rm(df_sdg, df_un, df_wb, maike_vars, other_vars)
# key files
# sdg raw data from UN SDG
# dat raw data from WorldBank

@@

```{r}
pca1 <- df_dat |> 
    select(-country_name) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> 
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

@@

```{r}
pca1$rotation |> 
    as_tibble(rownames = "vars") |> 
    separate(vars, sep = "_", into = c("year", "variable")) |> 
    mutate(year = as.numeric(year)) |> 
    pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "load") |> 
    mutate(PC = str_remove(PC, "PC"), PC = as.numeric(PC)) |> 
    filter(PC < 4) |> 
    ggplot(aes(year, load)) +
    geom_line(aes(color = variable)) +
    facet_wrap(~PC) +
    theme_light(base_size = 12)

@@

ggplot(data = x |> filter(var == "Proportion of Variance"), aes(comp, value)) +
    geom_col(color = "grey25") +
    # geom_point(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    # geom_line(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    labs(x = "Principal components", y = "Explained variance") +
    scale_x_continuous(n.breaks = 10) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_light(base_size = 6)

pca1$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_dat$country_code |> unique()) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    geom_segment(
        data = pca1$rotation |>
            as_tibble(rownames = "factor") |>
            separate(factor, into = c("year", "factor")),
        aes(xend = PC1, yend = PC2, x= 0, y = 0, color = factor, alpha = year)) +
    coord_equal()

@@

```{r}
pca2 <- df_un_reduced |> 
    unite(col = "var", series_name, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

@@

```{r}
mfa1 <- df_un_reduced |> 
    unite(col = "var", series_code, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 6), type = rep("s", 6)) # 18 years times 6 variables
```

@@

```{r}
ggbiplot::ggbiplot(
    pca1,
    labels = df_dat$country_code |> unique(),
    circle = TRUE,
    varname.abbrev = TRUE)
```

>>>>

countries_list <- countries |> filter(!is.na(currency_unit)) |> pull(country_code)
## correct Namibia, it is not an NA in x2_alpha_code

countries$x2_alpha_code[countries$country_code == "NAM"] <- "NA"

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Year", y = "Indicator name") +
    theme_light(base_size = 6)

@@

**J230323** Team recommended only to use inequality on UN Data which is available for ~50+ countries. Check newer dataset from WIID

```{r}
load("data/wii_small-data.RData")
```

```{r}
ineq_dat
```
`m` is total, `a` is average. All other codes are documented here: [https://wid.world/codes-dictionary/#packages](https://wid.world/codes-dictionary/#packages).

@@

```{r}
## both datasets are wider format, so makes sense to merge as is
df_dat <- df_dat |>
    left_join(countries |> select(country_code, country = x2_alpha_code)) |> 
    left_join(ineq_dat |> 
                  select(-age, -pop) |> 
                  unite(col = "var", c("variable", "percentile")) |> 
                  pivot_wider(names_from = var, values_from = value)
              )
```

@@

```{r}
b <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
     scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG14: Life under water", x = "Year", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6)
b
```

@@

## Very long names:
sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    select(indicator, indicator_name) |> 
    unique()

@@

Merge with WID inequality data:

df_un_reduced <- df_un_reduced |> ungroup() |> #group_by(iso3)
    left_join(
        countries |> select(iso3 = country_code, country = x2_alpha_code) ) |> 
    # filter(is.na(country)) |> pull(iso3) |> unique()
    unite(col = "var", year, series_name, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    left_join(
        ineq_dat |> 
            select(-age, -pop) |> 
            filter(year >=  2000, year < 2019) |> # max year 2018 in UN dataset
            unite(col = "series_name", c("variable", "percentile")) |> 
            right_join(
                countries |> 
                    select(iso3 = country_code, country = x2_alpha_code) |> 
                    filter(iso3 != "CHI") # filter out channel islands
            ) |> 
            unite(col = "var", year, series_name, sep = "_") |> 
            pivot_wider(values_from = value, names_from = var) 
        )
```

@@

```{r}
rm(df_sdg, df_un, df_wb, maike_vars, other_vars, ineq_dat)
# key files
# sdg raw data from UN SDG
# dat raw data from WorldBank
load("data/cleaned_sdg.RData")
```

@@

```{r}
pca1 <- df_dat |> 
    select(-country_name, -country, -sptinc992j_p0p100) |> 
    filter(!is.na(rptinc992j_p0p100)) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> 
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

@@

```{r}
p <- pca1$rotation |> 
    as_tibble(rownames = "vars") |> 
    separate(vars, sep = "_", into = c("year", "variable")) |> 
    mutate(year = as.numeric(year)) |> 
    pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "load") |> 
    mutate(PC = str_remove(PC, "PC"), PC = as.numeric(PC)) |> 
    filter(PC < 5) |> 
    ggplot(aes(year, load)) +
    geom_line(aes(color = variable)) +
    facet_wrap(~PC) +
    theme_light(base_size = 8)

# ggsave(
#     plot = p, file = "pca-wb-time.png", path = "figures/", device = "png",
#     width = 7, height = 4, bg = "white", dpi = 400
# )
p + scale_color_brewer(palette = "Set3")
```

@@

ggplot(data = x |> filter(var == "Proportion of Variance"), aes(comp, value)) +
    geom_col(color = "grey25") +
    geom_point(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    geom_line(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    labs(x = "Principal components", y = "Explained variance") +
    scale_x_continuous(n.breaks = 10) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_light(base_size = 6)

a <- pca1$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_dat |> 
               filter(!is.na(rptinc992j_p0p100)) |>
               pull(country_code) |> unique()) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    coord_equal() +
    theme_light(base_size = 6)

b <- ggplot() +
    geom_segment(
        data = pca1$rotation |>
            as_tibble(rownames = "factor") |> 
            mutate(year = str_sub(factor, start = 1L, end = 4L),
                   factor = str_remove(factor, pattern = "\\d{4}_"))  |> 
            select(year, factor, PC1:PC2),
        aes(xend = PC1, yend = PC2, color = factor, alpha = as.numeric(year)),
        x = 0, y = 0,
        arrow = arrow(length = unit(0.1,"cm")), linewidth = 0.3) +
    scale_color_brewer("Variable",palette = "Set3",
        guide = guide_legend(ncol = 2, title.position = "top")) +
    scale_alpha("Year", guide = guide_legend(ncol = 1, title.position = "top")) +
    coord_equal()+
    theme_light(base_size = 6)

a + b + plot_layout(guides = "collect") & 
    theme_light(base_size = 10) &
    theme(legend.position = "bottom")

# ggsave(
#     filename = "pca_wb-data.png", path = "figures/", dpi = 400, bg = "white",
#     device = "png", width = 7, height = 4,
#     plot = (a + b + plot_layout(guides = "collect") & 
#     theme_light(base_size = 6) &
#     theme(legend.position = "bottom") )
# )

@@

- 95% of the variability is explained by the first 10 PCs

For the UN dataset, the area based nature proxies have zero variance (SD), which means that despite completness, it is not informative.

```{r}
##
zero_variance_vars <- df_un_reduced |>
    pivot_longer(3:last_col(), names_to = "variable", values_to = "value") |> 
    mutate(
        year = str_sub(variable, 1L, 4L),
        variable = str_sub(variable, 6L, -1L)
    ) |> # pull(variable) |> unique()
    filter(variable != "sptinc992j_p0p100") |>
    group_by(iso3, variable) |> 
    mutate(variance = var(value)) |>
    filter(variance == 0) |>
    pull(variable) |> unique()

pca2 <- df_un_reduced |>
    ungroup() |> 
    select(-country, -iso3, -NA_NA) |> #skimr::skim()
    select(-contains("sptinc"), -contains("gptinc"), -contains("rptinc")) |> 
    prcomp(center = TRUE, scale. = TRUE)

```
**J230529:** It doesn't work, there is no enough variability in the data. It seems I have several variables / countries with zero variance or constants despite no missing values. However, it only screws up when adding the inequality variables. Without it does still run, why? Check with Maike. It might have to do with number of observations (countries) vs number of variables. Most likely too wide of a dataset can be problematic.

@@

```{r}
mfa1 <- df_un_reduced |> 
    select(-iso3, -country, -NA_NA) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 10), type = rep("s", 10)) # 18 years times 6 variables
```

@@

```{r}
ggbiplot::ggbiplot(
    pca1,
    labels = df_dat |>  select(-country_name, -country, -sptinc992j_p0p100) |> 
    filter(!is.na(rptinc992j_p0p100)) |> pull(country_code) |> unique(),
    circle = TRUE,
    varname.abbrev = TRUE)
```
